Think deep for reef resilience

### 2025-08-15 paper/paper_draft.qmd
Add section describing partial dependence analysis

We used a multinomial linear regression model to quantify the influence of reef
characteristics on reef cluster assignment. We model the influence of seven predictors,
median depth, log total connectivity strength, log source to sink ratio, mean DHW, log reef
area, mean DHW tolerance level and GCM on cluster assignment.

>>>>

### Partial Dependence Analysis

A Partial Dependence Analysis was conducted to understand the relationship between
individual features and time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest
including the median reef depth, mean DHW for the given scenario, total connectivity
strength, and the outgoing-to-incoming connectivity ratio (\$O:I\$). The full dataset
consisted of the timeseries classification of each of the 2429 reefs under the five GCMs
explored, for a total of 12,145 samples. The dataset was randomly split into training (60%)
and testing (40%) subsets to ensure model generalisability. The RF classifier achieved a 99%
accuracy on the training set and 66% on the test set.

While the 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 scores for Low, Medium, and High were determined to be 0.7, 0.57 and 0.68
respectively, indicating balanced performance across the cluster categories. Lower
performance with the Medium cluster reflect the inherent difficulty of classifying
transitional states in natural systems. For management applications, this level of
performance provides valuable decision support when combined with expert knowledge and the
model's probabilistic outputs.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual reef characteristic were systematically (one-at-a-time) varied across
a grid of 50 equally spaced points spanning its range, while all other features were left at
their modelled values. The procedure creates (1) a modified test dataset, where the target
reef characteristic was set to the grid value for all samples, (2) generates probabilistic
predictions for all three cluster classes using the trained RF classifier, and (3) the mean
predicted probability for each class was calculated across all test observations. The median
was also assessed (not shown) but was found not to affect the interpretation of the analysis
results. The process isolates the marginal effect of each feature by averaging over the
natural distribution of all other features in the test dataset.
